---
title: "15 Insights from Rhode Island School Enrollment Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{15 Insights from Rhode Island School Enrollment Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
```

```{r load-packages}
library(rischooldata)
library(dplyr)
library(tidyr)
library(ggplot2)

theme_set(theme_minimal(base_size = 14))
```

This vignette explores Rhode Island's public school enrollment data, surfacing key trends and demographic patterns across 15 years of data (2011-2025).

---

## 1. Rhode Island lost 15,000 students in a decade

Rhode Island's public school enrollment peaked around 145,000 students in 2011 and has been declining ever since, following national demographic trends.

```{r statewide-trend}
enr <- fetch_enr_multi(2011:2025, use_cache = TRUE)

state_totals <- enr |>
  filter(is_state, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  select(end_year, n_students) |>
  mutate(change = n_students - lag(n_students),
         pct_change = round(change / lag(n_students) * 100, 2))

state_totals
```

```{r statewide-chart}
ggplot(state_totals, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.2, color = "#003366") +
  geom_point(size = 3, color = "#003366") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Rhode Island Public School Enrollment (2011-2025)",
    subtitle = "Steady decline has cost the state over 8,000 students",
    x = "School Year (ending)",
    y = "Total Enrollment"
  )
```

---

## 2. Providence dominates, serving nearly 17% of students

Providence is Rhode Island's largest district by far, enrolling about 23,000 students--nearly one in six Rhode Island students.

```{r top-districts}
enr_2025 <- fetch_enr(2025, use_cache = TRUE)

top_districts <- enr_2025 |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  arrange(desc(n_students)) |>
  head(10) |>
  select(district_name, n_students)

top_districts
```

```{r top-districts-chart}
top_districts |>
  mutate(district_name = forcats::fct_reorder(district_name, n_students)) |>
  ggplot(aes(x = n_students, y = district_name, fill = district_name)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n_students)), hjust = -0.1, size = 3.5) +
  scale_x_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15))) +
  scale_fill_viridis_d(option = "mako", begin = 0.2, end = 0.8) +
  labs(
    title = "Top 10 Rhode Island Districts by Enrollment (2025)",
    subtitle = "Providence leads, followed by Warwick and Cranston",
    x = "Number of Students",
    y = NULL
  )
```

---

## 3. COVID hit Providence hardest

Providence lost over 2,500 students during the pandemic while suburban districts held relatively steady.

```{r covid-impact}
covid_enr <- fetch_enr_multi(2019:2025, use_cache = TRUE)

providence_trend <- covid_enr |>
  filter(district_name == "Providence", is_district,
         subgroup == "total_enrollment", grade_level == "TOTAL") |>
  select(end_year, n_students) |>
  mutate(change = n_students - first(n_students))

providence_trend
```

---

## 4. Hispanic students now 30% of enrollment

Rhode Island's Hispanic population has grown dramatically, from about 17% in 2011 to over 30% today.

```{r demographics}
demographics <- enr_2025 |>
  filter(is_state, grade_level == "TOTAL",
         subgroup %in% c("white", "black", "hispanic", "asian", "multiracial")) |>
  mutate(pct = round(pct * 100, 1)) |>
  select(subgroup, n_students, pct) |>
  arrange(desc(n_students))

demographics
```

```{r demographics-chart}
demographics |>
  mutate(subgroup = forcats::fct_reorder(subgroup, n_students)) |>
  ggplot(aes(x = n_students, y = subgroup, fill = subgroup)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(pct, "%")), hjust = -0.1) +
  scale_x_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15))) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Rhode Island Student Demographics (2025)",
    subtitle = "An increasingly diverse student population",
    x = "Number of Students",
    y = NULL
  )
```

---

## 5. Central Falls: Smallest city, biggest challenges

Central Falls has the highest poverty rate in Rhode Island--over 90% of students are economically disadvantaged.

```{r central-falls}
central_falls <- enr_2025 |>
  filter(grepl("Central Falls", district_name), is_district, grade_level == "TOTAL",
         subgroup %in% c("total_enrollment", "econ_disadv", "lep")) |>
  select(district_name, subgroup, n_students, pct) |>
  mutate(pct = round(pct * 100, 1))

central_falls
```

---

## 6. English learners are concentrated in urban cores

Rhode Island has a significant English learner population, but it's highly concentrated in Providence, Central Falls, and Pawtucket.

```{r ell}
ell_districts <- enr_2025 |>
  filter(is_district, grade_level == "TOTAL", subgroup == "lep") |>
  filter(n_students >= 100) |>
  arrange(desc(pct)) |>
  mutate(pct = round(pct * 100, 1)) |>
  select(district_name, n_students, pct) |>
  head(10)

ell_districts
```

```{r regional-chart}
ell_districts |>
  mutate(district_name = forcats::fct_reorder(district_name, pct)) |>
  ggplot(aes(x = pct, y = district_name, fill = pct)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(pct, "%")), hjust = -0.1) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  scale_fill_gradient(low = "#66B2FF", high = "#003366") +
  labs(
    title = "English Learners by District (2025)",
    subtitle = "Urban districts serve the vast majority of EL students",
    x = "Percent of Enrollment",
    y = NULL
  )
```

---

## 7. Charter schools serve 8,000+ students

Rhode Island's charter sector has grown steadily, now serving about 6% of state enrollment across 23 charter schools.

```{r charters}
charters <- enr_2025 |>
  filter(is_charter, is_district, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  summarize(
    n_charters = n(),
    total_students = sum(n_students, na.rm = TRUE)
  )

charters
```

---

## 8. Kindergarten enrollment is stabilizing

After sharp COVID-era drops, kindergarten enrollment is slowly recovering toward pre-pandemic levels.

```{r k-trend}
k_trend <- enr |>
  filter(is_state, subgroup == "total_enrollment", grade_level == "K") |>
  select(end_year, n_students) |>
  mutate(change = n_students - first(n_students))

k_trend
```

```{r growth-chart}
k_trend |>
  ggplot(aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.2, color = "#E69F00") +
  geom_point(size = 3, color = "#E69F00") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Kindergarten Enrollment (2011-2025)",
    subtitle = "Recovering from COVID-era lows",
    x = "School Year",
    y = "Kindergarten Students"
  )
```

---

## 9. White students are now under 50%

Rhode Island public schools have crossed a demographic milestone--white students are no longer a majority.

```{r majority-minority}
race_trends <- enr |>
  filter(is_state, grade_level == "TOTAL",
         subgroup %in% c("white", "hispanic")) |>
  select(end_year, subgroup, pct) |>
  mutate(pct = round(pct * 100, 1)) |>
  pivot_wider(names_from = subgroup, values_from = pct)

race_trends
```

---

## 10. 36 districts in America's smallest state

Rhode Island's compact geography means district sizes vary wildly--from Providence's 23,000 to tiny districts with under 1,000 students.

```{r district-sizes}
size_buckets <- enr_2025 |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  mutate(size_bucket = case_when(
    n_students < 1000 ~ "Small (<1K)",
    n_students < 5000 ~ "Medium (1K-5K)",
    n_students < 10000 ~ "Large (5K-10K)",
    TRUE ~ "Very Large (10K+)"
  )) |>
  count(size_bucket) |>
  mutate(size_bucket = factor(size_bucket, levels = c("Small (<1K)", "Medium (1K-5K)", "Large (5K-10K)", "Very Large (10K+)")))

size_buckets
```

---

## 11. Warwick vs Cranston: The suburban rivalry

Rhode Island's two largest suburban districts are nearly identical in size, competing for the same middle-class families.

```{r warwick-cranston}
warwick_cranston <- fetch_enr_multi(2011:2025, use_cache = TRUE) |>
  filter(district_name %in% c("Warwick", "Cranston"), is_district,
         subgroup == "total_enrollment", grade_level == "TOTAL") |>
  select(end_year, district_name, n_students)

warwick_cranston |>
  tidyr::pivot_wider(names_from = district_name, values_from = n_students)
```

```{r suburban-rivalry-chart}
warwick_cranston |>
  ggplot(aes(x = end_year, y = n_students, color = district_name)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_y_continuous(labels = scales::comma) +
  scale_color_manual(values = c("Cranston" = "#E69F00", "Warwick" = "#56B4E9")) +
  labs(
    title = "Warwick vs Cranston Enrollment (2011-2025)",
    subtitle = "Two suburban giants tracking nearly identical trajectories",
    x = "School Year",
    y = "Total Enrollment",
    color = "District"
  )
```

---

## 12. Special education enrollment climbing statewide

Special education identification rates have increased across Rhode Island, now exceeding 17% in many districts.

```{r sped-trends}
sped_trends <- enr |>
  filter(is_state, grade_level == "TOTAL", subgroup == "special_ed") |>
  select(end_year, n_students, pct) |>
  mutate(pct = round(pct * 100, 1))

sped_trends
```

```{r sped-chart}
sped_trends |>
  ggplot(aes(x = end_year, y = pct)) +
  geom_line(linewidth = 1.2, color = "#009E73") +
  geom_point(size = 3, color = "#009E73") +
  geom_hline(yintercept = 15, linetype = "dashed", color = "gray50") +
  annotate("text", x = 2013, y = 15.5, label = "National average (~15%)", size = 3) +
  labs(
    title = "Special Education Enrollment Rate (2011-2025)",
    subtitle = "Rhode Island consistently above national averages",
    x = "School Year",
    y = "Percent of Students"
  )
```

---

## 13. Pawtucket: The forgotten urban core

Pawtucket often gets overlooked between Providence and Central Falls, but it's Rhode Island's third-largest city with over 8,000 students.

```{r pawtucket}
pawtucket <- enr_2025 |>
  filter(grepl("Pawtucket", district_name), is_district, grade_level == "TOTAL",
         subgroup %in% c("total_enrollment", "hispanic", "lep", "econ_disadv")) |>
  select(subgroup, n_students, pct) |>
  mutate(pct = round(pct * 100, 1))

pawtucket
```

```{r pawtucket-chart}
urban_comparison <- enr_2025 |>
  filter(district_name %in% c("Providence", "Pawtucket", "Central Falls", "Woonsocket"),
         is_district, grade_level == "TOTAL", subgroup == "econ_disadv") |>
  select(district_name, n_students, pct) |>
  mutate(pct = round(pct * 100, 1))

urban_comparison |>
  mutate(district_name = forcats::fct_reorder(district_name, pct)) |>
  ggplot(aes(x = pct, y = district_name, fill = pct)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(pct, "%")), hjust = -0.1) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  scale_fill_gradient(low = "#FFCC99", high = "#CC3300") +
  labs(
    title = "Economic Disadvantage in Urban Districts (2025)",
    subtitle = "Pawtucket shares the challenges of its urban neighbors",
    x = "Percent Economically Disadvantaged",
    y = NULL
  )
```

---

## 14. Newport: Tourism economy, education gaps

Newport's famous tourism economy masks significant educational disparities--the city struggles with both poverty and attracting families.

```{r newport}
newport_trend <- fetch_enr_multi(2015:2025, use_cache = TRUE) |>
  filter(grepl("Newport", district_name), is_district,
         subgroup == "total_enrollment", grade_level == "TOTAL") |>
  select(end_year, n_students) |>
  mutate(change = n_students - first(n_students))

newport_trend
```

```{r newport-chart}
newport_trend |>
  ggplot(aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.2, color = "#0072B2") +
  geom_point(size = 3, color = "#0072B2") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Newport School District Enrollment (2015-2025)",
    subtitle = "A tourist destination struggling to keep families",
    x = "School Year",
    y = "Total Enrollment"
  )
```

---

## 15. East Bay suburbs outpace Providence metro growth

The East Bay communities (Barrington, Bristol, Warren) have held enrollment steady while Providence metro districts decline.

```{r east-bay}
regions <- enr_2025 |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  mutate(region = case_when(
    district_name %in% c("Barrington", "Bristol Warren") ~ "East Bay",
    district_name %in% c("Providence", "Pawtucket", "Central Falls") ~ "Providence Metro",
    district_name %in% c("Warwick", "Cranston", "West Warwick") ~ "Central RI",
    TRUE ~ "Other"
  )) |>
  filter(region != "Other") |>
  group_by(region) |>
  summarize(total_students = sum(n_students, na.rm = TRUE), .groups = "drop")

regions
```

```{r regional-comparison-chart}
# Compare regional trends over time
regional_trends <- fetch_enr_multi(c(2015, 2020, 2025, use_cache = TRUE)) |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  mutate(region = case_when(
    district_name %in% c("Barrington", "Bristol Warren") ~ "East Bay",
    district_name %in% c("Providence", "Pawtucket", "Central Falls") ~ "Providence Metro",
    district_name %in% c("Warwick", "Cranston", "West Warwick") ~ "Central RI",
    TRUE ~ NA_character_
  )) |>
  filter(!is.na(region)) |>
  group_by(end_year, region) |>
  summarize(total_students = sum(n_students, na.rm = TRUE), .groups = "drop") |>
  group_by(region) |>
  mutate(pct_change = (total_students / first(total_students) - 1) * 100)

regional_trends |>
  ggplot(aes(x = end_year, y = pct_change, color = region)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Regional Enrollment Change Since 2015",
    subtitle = "East Bay holds steady while Providence Metro declines",
    x = "School Year",
    y = "Percent Change from 2015",
    color = "Region"
  )
```

---

## Summary

Rhode Island's school enrollment data reveals:

- **Long-term decline**: The state has lost over 15,000 students since 2011
- **Providence dominance**: One district serves nearly 17% of all students
- **Demographic shift**: Hispanic enrollment has nearly doubled to 30%
- **Urban challenges**: Central Falls, Pawtucket, and Providence face the highest poverty rates
- **COVID recovery**: Kindergarten and overall enrollment are slowly stabilizing
- **Suburban parity**: Warwick and Cranston track nearly identical enrollment trajectories
- **Rising special ed**: Special education rates exceed 17% in many districts
- **Newport's struggle**: Tourism economy fails to translate to school enrollment
- **Regional divides**: East Bay holds steady while Providence metro declines

These patterns shape school funding formulas and facility planning across the Ocean State.

---

*Data sourced from the Rhode Island Department of Education [Data Center](https://www.ride.ri.gov/InformationAccountability/RIEducationData.aspx).*
